NumLin: Linear Types for Linear Algebra

### 2018-11-29 write-up/dissertation/background.tex
Organise write-ups
Add more evaluation examples

\begin{figure}[tp]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../test/examples/factorial.lt}
    \caption{Factorial function in LT4LA.}\label{fig:ltfla_factorial}
\end{figure}

\begin{figure}[tp]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../test/examples/sum_array.lt}
    \caption{Summing over an array in LT4LA.}\label{fig:ltfla_sumarray}
\end{figure}

@@

\begin{figure}[tp]
    \centering
    \begin{minted}[fontsize=\small]{ocaml}
let row = row[i] := x1 in (* or *) let () = free row in
(* Could not show equality:                                        *)
(*     z arr                                                       *)
(* with                                                            *)
(*     'x arr                                                      *)
(*                                                                 *)
(* Var 'x is universally quantified                                *)
(* Are you trying to write to/free/unshare an array you don't own? *)
(* In test/examples/sum_array.lt, at line: 7 and column: 19        *)
    \end{minted}
    \caption{Attempting to write to or free a read only array in
    LT4LA.}\label{fig:ltfla_univ}
\end{figure}

>>>>

\begin{figure}[tp]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../../examples/factorial.lt}
    \caption{Factorial function in LT4LA.}\label{fig:ltfla_factorial}
\end{figure}

\begin{figure}[tp]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../../examples/sum_array.lt}
    \caption{Summing over an array in LT4LA.}\label{fig:ltfla_sumarray}
\end{figure}

@@

\begin{figure}[tp]
    \centering
    \begin{minted}[fontsize=\small]{ocaml}
let row = row[i] := x1 in (* or *) let () = free row in
(* Could not show equality:                                        *)
(*     z arr                                                       *)
(* with                                                            *)
(*     'x arr                                                      *)
(*                                                                 *)
(* Var 'x is universally quantified                                *)
(* Are you trying to write to/free/unshare an array you don't own? *)
(* In examples/sum_array.lt, at line: 7 and column: 19        *)
    \end{minted}
    \caption{Attempting to write to or free a read only array in
    LT4LA.}\label{fig:ltfla_univ}
\end{figure}

### 2018-11-29 write-up/paper/appendix.tex
Finish Performance Metrics (for Kalman, for now)
Add more evaluation examples

\vspace*{\fill}
\begin{figure}[h]
\begin{center}
    \begin{minted}[fontsize=\footnotesize]{ocaml}
let kalman sigma h mu r_1 data_1 =
  let h, _p_k_n_p_ = Prim.size_mat h in
  let k, n = _p_k_n_p_ in
  let sigma_hT = Prim.matrix n k in
  let (sigma, h), sigma_hT =
    Prim.gemm (Many 1.) (sigma, Many false) (h, Many true) (Many 0.) sigma_hT in
  let (h, sigma_hT), r_2 =
    Prim.gemm (Many 1.) (h, Many false) (sigma_hT, Many false) (Many 1.) r_1 in
  let k_by_k, x = Prim.posv_flip r_2 sigma_hT in
  let (h, mu), data_2 =
    Prim.gemm (Many 1.) (h, Many false) (mu, Many false) (Many (-1.)) data_1 in
  let (x, data_2), new_mu =
    Prim.gemm (Many 1.) (x, Many false) (data_2, Many false) (Many 1.) mu in
  let x_h = Prim.matrix n n in
  let (x, h), x_h =
    Prim.gemm (Many 1.) (x, Many false) (h, Many false) (Many 0.) x_h in
  let () = Prim.free_mat x in
  let sigma, sigma2 = Prim.copy_mat sigma in
  let (sigma, x_h), new_sigma =
    Prim.symm (Many true) (Many (-1.)) sigma x_h (Many 1.) sigma2 in
  let () = Prim.free_mat x_h in
  ((sigma, h), (new_sigma, (new_mu, (k_by_k, data_2))))
    \end{minted}
    \caption{OCaml code for a Kalman filter, generated (at \emph{compile time})
        from the code in Figure~\ref{fig:lang_kalman}, passed through
        \texttt{ocamlformat} for presentation.}\label{fig:ocaml_kalman}
\end{center}
\end{figure}
\vfill

\begin{landscape}
\begin{figure}[p]
    \centering
    \begin{minted}[fontsize=\footnotesize]{c}
static void kalman( const int n,               const int k,                const double *sigma, /* n,n */
                    const double *h, /* k,n */ const double *mu,  /* n,1 */ double *r,           /* k,k */
                    double *data,    /* k,1 */ double **ret_sigma /* n,n */ ) {
    double* n_by_k = (double *) malloc(n * k * sizeof(double));
    cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasTrans, n, k, n, 1., sigma, n, h, n, 0., n_by_k, k);
    cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasNoTrans, k, k, n, 1., h, n, n_by_k, k, 1., r, k);
    LAPACKE_dposv(LAPACK_COL_MAJOR, 'U', k, n, r, k, n_by_k, k);
    cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasNoTrans, k, 1, n, 1., h, n, mu, 1, -1., data, 1);
    cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasNoTrans, n, 1, k, 1., n_by_k, k, data, 1, 1., mu, 1);
    double* n_by_n = (double *) malloc(n * n * sizeof(double));
    cblas_dgemm(CblasRowMajor, CblasNoTrans, CblasNoTrans, n, n, k, 1., n_by_k, k, h, n, 0., n_by_n, n);
    free(n_by_k);
    double* n_by_n2 = (double *) malloc(n * n * sizeof(double));
    cblas_dcopy(n*n, sigma, 1, n_by_n2, 1);
    cblas_dsymm(CblasRowMajor, CblasRight, CblasUpper, n, n, -1., sigma, n, n_by_n, n, 1., n_by_n2, n);
    free(n_by_n);
    *ret_sigma = n_by_n2; }
    \end{minted}
    \caption{CBLAS/LAPACKE implementation of a Kalman filter. I used C instead
        of Fortran because it is what Owl uses under the hood and OCaml FFI
        support for C is better and easier to use than that for Fortran. A distinct
        `measure\_kalman' function that sandwiches a call to this function with
        \texttt{getrusage} is omitted for brevity.}\label{fig:cblas_kalman}

\end{figure}
\end{landscape}

>>>>

\begin{figure}[h]
\begin{center}
    \begin{minted}[fontsize=\footnotesize]{ocaml}
let kalman sigma h mu r_1 data_1 =
  let h, _p_k_n_p_ = Prim.size_mat h in
  let k, n = _p_k_n_p_ in
  let sigma_hT = Prim.matrix n k in
  let (sigma, h), sigma_hT =
    Prim.gemm (Many 1.) (sigma, Many false) (h, Many true) (Many 0.) sigma_hT in
  let (h, sigma_hT), r_2 =
    Prim.gemm (Many 1.) (h, Many false) (sigma_hT, Many false) (Many 1.) r_1 in
  let k_by_k, x = Prim.posv_flip r_2 sigma_hT in
  let (h, mu), data_2 =
    Prim.gemm (Many 1.) (h, Many false) (mu, Many false) (Many (-1.)) data_1 in
  let (x, data_2), new_mu =
    Prim.gemm (Many 1.) (x, Many false) (data_2, Many false) (Many 1.) mu in
  let x_h = Prim.matrix n n in
  let (x, h), x_h =
    Prim.gemm (Many 1.) (x, Many false) (h, Many false) (Many 0.) x_h in
  let () = Prim.free_mat x in
  let sigma, sigma2 = Prim.copy_mat sigma in
  let (sigma, x_h), new_sigma =
    Prim.symm (Many true) (Many (-1.)) sigma x_h (Many 1.) sigma2 in
  let () = Prim.free_mat x_h in
  ((sigma, h), (new_sigma, (new_mu, (k_by_k, data_2))))
    \end{minted}
    \caption{OCaml code for a Kalman filter, generated (at \emph{compile time})
        from the code in Figure~\ref{fig:lang_kalman}, passed through
        \texttt{ocamlformat} for presentation.}\label{fig:ocaml_kalman}

\vspace{\baselineskip}

    \begin{minted}[fontsize=\footnotesize]{c}
static void kalman( const int n,                   const int k,                
                    const double *sigma, /* n,n */ const double *h,   /* k,n */
                    const double *mu,    /* n,1 */ double *r,         /* k,k */
                    double *data,        /* k,1 */ double **ret_sigma /* n,n */ ) {
    double* n_by_k = (double *) malloc(n * k * sizeof(double));
    cblas_dgemm(RowMajor, NoTrans, Trans, n, k, n, 1., sigma, n, h, n, 0., n_by_k, k);
    cblas_dgemm(RowMajor, NoTrans, NoTrans, k, k, n, 1., h, n, n_by_k, k, 1., r, k);
    LAPACKE_dposv(LAPACK_COL_MAJOR, 'U', k, n, r, k, n_by_k, k);
    cblas_dgemm(RowMajor, NoTrans, NoTrans, k, 1, n, 1., h, n, mu, 1, -1., data, 1);
    cblas_dgemm(RowMajor, NoTrans, NoTrans, n, 1, k, 1., n_by_k, k, data, 1, 1., mu, 1);
    double* n_by_n = (double *) malloc(n * n * sizeof(double));
    cblas_dgemm(RowMajor, NoTrans, NoTrans, n, n, k, 1., n_by_k, k, h, n, 0., n_by_n, n);
    free(n_by_k);
    double* n_by_n2 = (double *) malloc(n * n * sizeof(double));
    cblas_dcopy(n*n, sigma, 1, n_by_n2, 1);
    cblas_dsymm(RowMajor, Right, Upper, n, n, -1., sigma, n, n_by_n, n, 1., n_by_n2, n);
    free(n_by_n);
    *ret_sigma = n_by_n2; }
    \end{minted}
    \caption{CBLAS/LAPACKE implementation of a Kalman filter. I used C instead
        of Fortran because it is what Owl uses under the hood and OCaml FFI
        support for C is better and easier to use than that for Fortran. A distinct
        `measure\_kalman' function that sandwiches a call to this function with
        \texttt{getrusage} is omitted for brevity.}\label{fig:cblas_kalman}

\end{center}
\end{figure}

### 2018-11-29 write-up/paper/implementation.tex
Add more evaluation examples

For performance, we measured the execution of four equivalent
implementations of a Kalman filter: in C (using \textsc{Cblas}), \lang\ (using
\textsc{Owl}'s low-level \textsc{Cblas} bindings), OCaml (using \textsc{Owl}'s
intended, safe/copying-by-default interface), and Python (using \textsc{NumPy},
with the interpreter started and functions interpreted). We measured execution
time in micro-seconds, against an exponentially (powers of 5) increasing
scaling factor for matrix size parameters $n=5$ and $k=3$.

@@

@@

The results in Figure~\ref{fig:timings} are as we expected: C is the fastest,
followed by \lang, with OCaml and Python last.  Differences in timings are
quite pronounced at small matrix sizes, but are still significant at larger
ones: for $n=625$, \textsc{Cblas} took $112 \pm 35\, ms$, \lang\ took $105 \pm
25 \, ms$, \textsc{Owl} took $124 \pm 38\, ms$ and \textsc{NumPy} took $112
\pm 12\, ms$;  for $n=3125$, \textsc{Cblas} took $10.8 \pm 0.7\, s$, \lang\
took $12.0 \pm 1.2 \, s$, \textsc{Owl} took $13.3 \pm 0.2\, s$ and
\textsc{NumPy} took $12.7 \pm 0.6\, s$.

Worth highlighting here is the other major advantage of using \lang\ is reduced
memory usage.  Whilst the \textsc{Owl} and \textsc{NumPy} use 11 temporary
matrices (\emph{excluding} the 2 matrices which store the results), using $n +
n^2 + 4nk + 3k^2 + 2k \approx 4n^2$ (for $k = 3n/5$) words of memory,
\textsc{Cblas} and \lang\ use only \emph{2} temporary m`atrices (excluding the
\emph{one} matrix which stores one of the results), using only $n^2 + nk \leq
2n^2$ words of memory.

\begin{figure}[t]
    \centering
    \input{timings.tex}
    \caption{Comparison of execution times (error bars are present but quite
        small). Small matrices and timings $n \le 5^3$ were micro-benchmarked
        with the Core\_bench library. Larger ones used Unix's
        \texttt{getrusage} functionality, sandwiched between calls to
        \highl{Gc.full_major} for the OCaml implementations.}\label{fig:timings}
\end{figure}

@@

% Other examples impl
% Sampled linear reg example

>>>>

For performance, we measured the execution times of four equivalent
implementations of a Kalman filter: in C (using \textsc{Cblas}), \lang\ (using
\textsc{Owl}'s low-level \textsc{Cblas} bindings), OCaml (using \textsc{Owl}'s
intended, safe/copying-by-default interface), and Python (using \textsc{NumPy},
with the interpreter started and functions interpreted). We measured execution
time in micro-seconds, against an exponentially (powers of 5) increasing
scaling factor for matrix size parameters $n=5$ and $k=3$.

@@

We also measured the execution times of L1-norm minimisation and the
``linear-regression'' ($\mathbf{(X^T X)^{-1} X^T y}$) similary, but without a C
implementation.

@@

The results in Figures~\ref{fig:timings} are as we expected: C is the fastest,
followed by \lang, with OCaml and Python last.  Differences in timings are
quite pronounced at small matrix sizes, but are still significant at larger
ones. Specifically for the Kalman filter, for $n=625$, \textsc{Cblas} took $112
\pm 35\, ms$, \lang\ took $105 \pm 25 \, ms$, \textsc{Owl} took $124 \pm 38\,
ms$ and \textsc{NumPy} took $112 \pm 12\, ms$;  for $n=3125$, \textsc{Cblas}
took $10.8 \pm 0.7\, s$, \lang\ took $12.0 \pm 1.2 \, s$, \textsc{Owl} took
$13.3 \pm 0.2\, s$ and \textsc{NumPy} took $12.7 \pm 0.6\, s$.

Worth highlighting here is the other major advantage of using \lang\ is reduced
memory usage.  Whilst the \textsc{Owl} and \textsc{NumPy} use 11 temporary
matrices for the Kalman filter, (\emph{excluding} the 2 matrices which store
the results), using $n + n^2 + 4nk + 3k^2 + 2k \approx 4n^2$ (for $k = 3n/5$)
words of memory, \textsc{Cblas} and \lang\ use only \emph{2} temporary
matrices (excluding the \emph{one} matrix which stores one of the results),
using only $n^2 + nk \leq 2n^2$ words of memory.

\begin{figure}[t]
    \centering
    \input{kalman_timings.tex}
    \begin{minipage}{.49\textwidth}
        \centering
        \input{l1_norm_min_timings.tex}
    \end{minipage}
    \begin{minipage}{.49\textwidth}
        \centering
        \input{lin_reg_timings.tex}
    \end{minipage}
    \caption{Comparison of execution times (error bars are present but quite
        small). Small matrices and timings $n \le 5^3$ were micro-benchmarked
        with the Core\_bench library. Larger ones used Unix's
        \texttt{getrusage} functionality, sandwiched between calls to
        \highl{Gc.full_major} for the OCaml implementations.}\label{fig:timings}
\end{figure}

@@

### 2018-11-29 write-up/paper/lang_and_examples.tex
Add more evaluation examples

\begin{figure}[t]
    \centering
    \inputminted[fontsize=\small, linenos]{ocaml}{../../test/examples/factorial.lt}
    \caption{Factorial function in \lang.}\label{fig:lang_factorial}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../../test/examples/sum_array.lt}
    \caption{Summing over an array in \lang.}\label{fig:lang_sumarray}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \begin{minted}[fontsize=\small]{ocaml}
let row = row[i] := x1 in (* or *) let () = free row in
(* Could not show equality:                                        *)
(*     z arr                                                       *)
(* with                                                            *)
(*     'x arr                                                      *)
(*                                                                 *)
(* Var 'x is universally quantified                                *)
(* Are you trying to write to/free/unshare an array you don't own? *)
(* In test/examples/sum_array.lt, at line: 7 and column: 19        *)
    \end{minted}
    \caption{Attempting to write to or free a read only array in
    \lang.}\label{fig:lang_errormsg}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../../test/examples/weighted_avg_infer.lt}
    \caption{\emph{Simplified} one-dimensional convolution.}\label{fig:lang_oned_conv}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../../test/examples/square.lt}
    \caption{Linear regression (OLS): $\hat\beta =
        \mathbf{(X^T X)^{-1} X^T y}$}\label{fig:lang_square}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../../test/examples/lin_reg.lt}
    \caption{Linear regression (OLS): $\hat\beta =
        \mathbf{(X^T X)^{-1} X^T y}$}\label{fig:lang_lin_reg}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../../test/examples/l1_norm_min.lt}
    \caption{L1-norm minimisation on manifolds:
        $\mathbf{Q^{-1}U(I+U^TQ^{-1}U)^{-1}U^T}$}\label{fig:lang_l1_norm_min}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../../test/examples/kalman.lt}
    \caption{Kalman filter: see Figure~\ref{fig:kalman_eqns} for the
        equations this code implements. Line numbers in comments refer to
        equivalent lines in a C implementation
        (Figure~\ref{fig:cblas_kalman}).}\label{fig:lang_kalman}
\end{figure}

>>>>

\begin{figure}[t]
    \centering
    \inputminted[fontsize=\small, linenos]{ocaml}{../../examples/factorial.lt}
    \caption{Factorial function in \lang.}\label{fig:lang_factorial}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../../examples/sum_array.lt}
    \caption{Summing over an array in \lang.}\label{fig:lang_sumarray}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \begin{minted}[fontsize=\small]{ocaml}
let row = row[i] := x1 in (* or *) let () = free row in
(* Could not show equality:                                        *)
(*     z arr                                                       *)
(* with                                                            *)
(*     'x arr                                                      *)
(*                                                                 *)
(* Var 'x is universally quantified                                *)
(* Are you trying to write to/free/unshare an array you don't own? *)
(* In examples/sum_array.lt, at line: 7 and column: 19        *)
    \end{minted}
    \caption{Attempting to write to or free a read only array in
    \lang.}\label{fig:lang_errormsg}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../../examples/weighted_avg_infer.lt}
    \caption{\emph{Simplified} one-dimensional convolution.}\label{fig:lang_oned_conv}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../../examples/square.lt}
    \caption{Linear regression (OLS): $\hat\beta =
        \mathbf{(X^T X)^{-1} X^T y}$}\label{fig:lang_square}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../../examples/lin_reg.lt}
    \caption{Linear regression (OLS): $\hat\beta =
        \mathbf{(X^T X)^{-1} X^T y}$}\label{fig:lang_lin_reg}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../../examples/l1_norm_min.lt}
    \caption{L1-norm minimisation on manifolds:
        $\mathbf{Q^{-1}U(I+U^TQ^{-1}U)^{-1}U^T}$}\label{fig:lang_l1_norm_min}
\end{figure}

@@

\begin{figure}[t]
    \centering
%   \begin{minipage}{.4\textwidth}
%   \centering
    \inputminted[fontsize=\small]{ocaml}{../../examples/kalman.lt}
%   \end{minipage}
%   \begin{minipage}{.4\textwidth}
%   \centering
%   \begin{minted}[linenos,fontsize=\small]{ocaml}
%   let owl_kalman
%       ~sigma
%       ~h
%       ~mu
%       ~r
%       ~data =
%     let open Owl.Mat in
%     let ( * ) = dot in
%     let h' = transpose h in
%     let sigma_h' = sigma * h' in
%     let x = sigma_h' * (inv @@ r + h * sigma_h') in
%     let new_mu = mu + x * (h * mu - data) in
%     let new_sigma = sigma - x * h * sigma in
%     new_sigma, new_mu
%   ;;
%   \end{minted}
%   \end{minipage}
    \caption{Kalman filter: see Figure~\ref{fig:kalman_eqns} for the
        equations this code implements. Line numbers in comments refer to
        equivalent lines in a C implementation
        (Figure~\ref{fig:cblas_kalman}).}\label{fig:lang_kalman}
\end{figure}
